---
title: "Empirical Economics"
subtitle: "Lecture 5: Panel Data II"
format:
  revealjs: 
    theme: [default, theme.scss]
    slide-number: true
    chalkboard: 
      buttons: false
    preview-links: auto
    logo: logo.svg
    css: styles.css
    footer: 'Empirical Economics: Lecture 5 - Panel Data II'
---

```{r setup}
#| warning: false
#| message: false
#| echo: false

library(ggplot2)
library(gridExtra)
library(here)
library(reticulate)
python_path <- here::here(".venv", "bin", "python")
use_python(python_path)
```

# Outline

## Course Overview

- Linear Model I
- Linear Model II
- Time Series and Prediction
- Panel Data I
- Panel Data II
- Binary Outcome Data
- Potential Outcomes and Difference-in-differences
- Instrumental Variables


## This lecture

- Classification of panel data models
- Pooled OLS estimator
- Random-effects estimator
- Random effects or fixed effects: Hausman test
- Pooled OLS or Random effects
- Line of reasoning with panel data

- **Material:** Wooldridge Chapter 13: 13.1, 13.3, 13.4, 13.5

## Motivation

- Example: panel information of 900 international businesses (firms). Years: 2019, 2020, 2021.
- Variable of interest: Profits of these international firms.
- Research question: can we explain the volume of profits by the country of origin (= location of headquarter)?
  - E.g., firm has its headquarter in UK, The Netherlands, Germany or France (4 options).

- For one year (cross section)? Answer: YES (reference group: Germany)

  $$
    \text{Profits}_{i} = \beta_{0} + \beta_{1}UK_{i} + \beta_{2}France_{i} + \beta_{3}Netherlands_{i} + u_{i}
  $$
  - for $i=1,...,900$
  
- For three years? Answer: It depends.
  - This week: various panel methods contrasted.

## Motivation (Cont.)

- Regression equation is specified as:

  $$
    \text{Profits}_{it} = a_{i} + \beta_{1}UK_{i} + \beta_{2}France_{i} + \beta_{3}Netherlands_{i} + u_{it}
  $$
  
  - with $i = 1,...,900; t = 2019,2020,2021$

- UK, France, Netherlands and Germany are 0-1 indicators: 
  
  $$
    UK_{i} + France_{i} + Netherlands_{i} + Germany_{i} = 1
  $$

## Motivation (Cont.)

- Can we estimate the regression parameters by the first difference estimator? 
  - Answer: No, because $\Delta UK_{i} = 0$, $\Delta France_{i} = 0$ and $\Delta Netherlands_{i} = 0$
  
  $$
    \Delta \text{Profits}_{it} = \beta_{1}\Delta UK_{i} + \beta_{2}\Delta France_{i} + \beta_{3}\Delta Netherlands_{i} + \Delta u_{it}
  $$

- Can we estimate the regression parameters with the fixed effects estimator?
  - Answer: No, because $\ddot{UK}_{it} = UK_{it} - \tfrac{1}{T}\sum_{t=1}^Y UK_{it} = 0$


- Can we estimate the parameters by the **Pooled OLS estimator**?
  - Answer: Yes:
  
  $$
    \text{Profits}_{it} = \beta_{0} + \beta_{1}UK_{i} + \beta_{2}France_{i} + \beta_{3}Netherlands_{i} + v_{it}
  $$
  
- with $v_{it} = a_{i} + u_{it}$, $i = 1,...,900;t = 2019,2020,2021$

## Three Questions

- **Question 1:** Can we apply alternative estimators, next to the Fixed Effects and the First Difference estimator? This week: Pooled OLS.

- **Question 2:** Can we have a panel data estimator that includes country of headquarter? This week: **Random Effects estimator**

- **Question 3:** Is there any road map for the preferred estimator? 
  - Five options: 1. First Difference estimator, 2. LSDV estimator, 3. Fixed Effects Estimator, 4. Pooled OLS, 5. Random Effects estimator. 
  - Also this week: several testing procedures.
  
# Classification of Panel Models

## Static Model: A Classification

- Consider the following equation:

  $$
    y_{it} = a_{i} + \beta_{1}x_{1it} + ... + \beta_{k}x_{kit} + u_{it} \quad i=1,...,N; t=1,...,T
  $$

- It leads to the following two questions:
- **Question 1:** Is there any nonzero correlation between $a_{i}$ and all of the k RHS vars $x_{it}$? Is it nonzero in all T time periods?
  - $E(a_{i} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}) = 0$ 
  - $E(a_{i} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}) \neq 0$?
   
## Static Model: A Classification (Cont.)

- **Question 2:** Are all of the k variables $x_{it}$ strictly exogenous (conditional on the unobserved individual effect $a_{i}$)?
  - $E(u_{it} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}, a_{i}) = 0$ (thus, no lagged dependent variables, no feedback mechanism)
- In all of the estimators of last week we assumed that the exogenous time-varying regressors are strictly exogenous. (conditional on the unobserved effect):
  - $E(u_{it} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}, a_{i}) = 0$
  - So again: no lagged dependent variables; no feedback mechanism.
  - In explaining estimation procedures, we assume a **balanced panel**: for every cross-sectional unit, we have the same number of time periods *T*.

- Model: $y_{it} = a_{i} + \beta_{1}x_{1it} + ... + \beta_{k}x_{kit} + u_{it} \quad i=1,...,N;t=1,...,T$
  - Software allows for estimation of **unbalanced panels**, in which not all units have *T* observations. Thus the i-th individual has $T_{i}$ observations.


## Table A: Estimation Methods

- Estimation methods under different assumptions of strict exogeneity and on the correlation between the individual effect and RHS-variables:

:::{style="font-size: 1.2em;"}

| | $E(a_{i} | X_{i11},..,X_{i1k},..,X_{iT1},..,X_{iTk}) \neq 0$, Correlation between $a_{i}$ and all of the explanatory variables is allowed to be nonzero | $E(a_{i} | X_{i11},..,X_{i1k},..,X_{iT1},..,X_{iTk}) = 0$, A zero correlation between $a_{i}$ and all of the explanatory variables is assumed. |
| :--- | :--- | :--- |
| **All $x_{it}$ strictly exogenous** | 1. First differences <br> 2. LSDV procedure <br> 3. Within estimation | 4. Random effects |
| **Some $x_{it}$ not strictly exogenous** | Instrumental Variables (IV) | 5. Pooled OLS (no lagged dependent variables) <br> 6. Instrumental variables (IV) (lagged dep. vars. Included) |

:::


## Table B: Estimating Effect of Time-Invariant Variables

:::{style="font-size: 1.2em;"}

| | No estimation of $\gamma$: no effect of $z_{i}$ on $y_{it}$ | Estimation of $\gamma$: effect of $z_{i}$ on $y_{it}$ |
| :--- | :--- | :--- |
| **All $x_{it}$ strictly exogenous** | 1. First differences<br>2. LSDV procedure<br>3. Within estimation | 4. Random effects |
| **Some $x_{it}$ not strictly exogenous** | Instrumental Variables (IV) | 5. Pooled OLS (no lagged dependent variables)<br>6. Instrumental variables (IV) (lagged dep. vars. Included) |

::: 

- The model here is: $y_{it} = a_{i} + \beta_{1}x_{1it} + ... + \beta_{k}x_{kit} + \gamma_{1}z_{1i} + ... + \gamma_{h}z_{hi} + u_{it}$
  *   The variables $z$ do not change across time but they are different across individuals
  *   E.g. $z_{i}$: gender and ethnicity in wage equation
  *   $z_{i}$ picks up the between variation (between individuals).
      *   Between individuals: cross-sectional perspective
  *   $x_{it}$ picks up the within variation (within individuals)
      *   Within individuals: time-series perspective for a given individual
  *   Economists are usually more interested in within variation.

# Pooled OLS

## Assumptions 

- Suppose that instead one estimates the following model by OLS:
    $$y_{it} = \beta_1x_{1it} + ... + \beta_kx_{kit} + v_{it} \quad i=1,\dots,N;t=1,\dots,T$$

- Where $v_{it} = a_i + u_{it}$ (=individual specific effect + idiosyncratic error term)

- In this model, is it assumed that:
- $E(v_{it} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}) = 0$ so that, in all $k$ explanatory variables in all $T$ time periods:
  - $E(a_i | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}) = 0$ and 
  - $E(u_{it} | x_{i11},..,x_{i1k},..,x_{iT1},...,x_{iTk}, a_i) = 0$ 

- Violation of this assumption leads to a biased estimate of the regression parameters $\beta$.
  - An application will be given later.


## Strict Exogeneity

- In simple words strict exogeneity means we have:
  - No models with a lagged dependent variable
  - No models with a feedback mechanism: $y(t) \leftarrow x(t) \leftarrow y(t-1) \leftarrow x(t-1)$

:::{.callout-note title="Assumption TS.2 (Strict exogeneity)"}

For each $t$, the expected value of $u_t$ given ALL of the $k$ explanatory variables FOR ALL $T$ time periods, is equal to zero: $E[u_t | X] = 0$
  
:::

:::{.callout-note title="Assumption TS.2 (Contemporaneous exogeneity)"}

For each *t*, the expected value of $u_t$, given ALL of the *k* explanatory variables in period *t*, is equal to zero: $E(u_t | x_{t1},...,x_{tk}) = E(u_t | \mathbf{x}_t) = 0$

This assumption implies that the error term in period *t* is uncorrelated with all *k* regressors in the same period, $t$: $Corr(u_t,x_{tj})=0 \quad j=1,...,k$

:::


## Violation of Exogeneity

- Violation of the strict exogeneity assumption while assumption contemporaneous exogeneity is satisfied

:::{.callout-tip title="Example: Dynamic variable with lagged dependent variable"}

$y_t = \alpha_0 + \alpha_1y_{t-1} + \delta_1z_t + u_t$

- $u_t$ is assumed to be an idiosyncratic error term and contemporaneously exogenous: $E(u_t | y_{t-1}, z_t) = 0$, so that OLS yields consistent estimates.
- However, $y_{t-1}$ is NOT a strictly exogenous variable:
  - The assumption of strict exogeneity implies that $u_t$ is uncorrelated not only with $\mathbf{x}_t = (y_{t-1}, z_t)'$, but also with $\mathbf{x}_{t+1} = (y_t, z_{t+1})'$
  - According to equation (7), $y_t$ and $u_t$ are related to each other. In other words, $E(u_t | \mathbf{x}_{t+1}) = E(u_t | y_t, z_{t+1}) \neq 0$,
- Thus the lagged dependent variable $y_{t-1}$ is not strictly exogenous the model.

:::

## Violation of Exogeneity Examples

:::{.callout-tip title="Example: Models with a feedback mechanism"}

$gGDP_t = \alpha_0 + \delta_0r_t + u_t$ (8)

*   gGDP: GDP-growth rate
*   $r_t$: Interest rate, which is assumed to be contemporaneously exogenous: $E(u_t | r_t) = 0$
*   The independent variable $r_t$ depends on the lagged value of the dependent variable (feedback mechanism):
    $r_t = \gamma_0 + \gamma_1(gGDP_{t-1} – 3) + v_t$ (9)
*   Equation (9) implies that $r_{t+1}$ depends on $gGDP_t$ and consequently on $u_t$.
*   $E(u_t | \mathbf{x}_{t+1}) = E(u_t | r_{t+1}) \neq 0$
*   **Thus $r_t$ is not strictly exogenous in the model.**

:::


## Pooled OLS Estimator

- The Pooled OLS Estimator is an estimator for a zero correlation between $\alpha_i$ and the explanatory variables

- Now we assume that there is no correlation between $\alpha_i$ and $x_{it}$
- One of two methods we treat this week:
  - Pooled OLS (now)
  - Random effects (later)

## Pooled OLS Estimation

*   Now we assume that there is no correlation between $a_i$ and all of the $k$ explanatory variables $x_{it}$.
*   Equation (12) can be estimated with OLS. This is referred to as pooled OLS:
    $$y_{it} = \beta_1x_{1it} + \dots + \beta_kx_{kit} + v_{it} \quad (12)$$
    *   For which $v_{it} = a_i + u_{it}$ where $u_{it}$ is i.i.d.
    *   As a result of the identical distribution: $\text{Var}(v_{it}) = \text{Var}(v_{i,t-1})$
    *   In a pooled regression $v_{it} = a_i + u_{it}$
    *   So, in equation (12), it is assumed that:
        $E(a_i | x_{i11}, \dots, x_{i1k}, \dots, x_{it1}, \dots, x_{itk}) = 0$
        and
        $E(u_{it} | x_{i11}, \dots, x_{i1k}, \dots, x_{it1}, \dots, x_{itk}, a_i) = 0$
*   $a_i$ is uncorrelated with all of the explanatory variables.
*   $u_{it}$ is uncorrelated with all of the explanatory variables and $a_i$.

## Autocorrelation in Pooled OLS

- The autocorrelation between the error terms $v_{it}$ and $v_{i,t-1}$ of equation (12) is:

  $$
  \text{Corr}(v_{it}, v_{i,t-1}) = \frac{\text{Cov}(v_{it}, v_{i,t-1})}{\sqrt{\text{Var}(v_{it})\text{Var}(v_{i,t-1})}} = \frac{\text{Cov}(v_{it}, v_{i,t-1})}{\text{Var}(v_{it})}
  $$

- We can show that the numerator is:

  $$
  \begin{align*}
  \text{Cov}(v_{it}, v_{i,t-1}) &= \text{Cov}(a_i + u_{it}, a_i + u_{i,t-1}) \\
  &= \text{Cov}(a_i, a_i) + \text{Cov}(a_i, u_{i,t-1}) + \text{Cov}(u_{it}, a_i) + \text{Cov}(u_{it}, u_{i,t-1}) \\
  &= \sigma_a^2 \quad (=0) \quad (=0) \quad (=0) \\
  &= \text{Var}(a_i) + 0 + 0 + 0 \\
  &= \sigma_a^2
  \end{align*}
  $$

## Autocorrelation in Pooled OLS (Cont.)

- And the denominator is:
$$
\begin{align*}
\text{Var}(v_{it}) &= \text{Var}(a_i + u_{it}) \\
&= \text{Var}(a_i) + \text{Var}(u_{it}) + 2\text{Cov}(a_i, u_{it}) \\
&= \sigma_a^2 + \sigma_u^2 \quad (=0) \\
&= \sigma_a^2 + \sigma_u^2
\end{align*}
$$ 

- So that:

  $$
  \frac{\text{Cov}(v_{it}, v_{i,t-1})}{\text{Var}(v_{it})} = \frac{\sigma_a^2}{\sigma_a^2 + \sigma_u^2}
  $$

- Conclusion: In pooled OLS there is always autocorrelation.
  - The estimation procedure for Pooled OLS: OLS on with Newey-West robust standard errors, which are similar to clustered standard errors. It corrects for both heteroskedasticity and autocorrelation.

## Example: Pooled OLS

- Example: Vella and Verbeek (1998) ask whether unions raise wages. 
  - They have a panel dataset with individual wages, union membership, and various control variables, of which some are time-varying (e.g. experience), and some are constant over time (e.g. race).

- **Estimation procedure:**
  *   We start with the pooled OLS estimator.
  *   Next, we check for autocorrelation, using the Breusch Godfrey test.
  *   The parameter on the lagged residual is statistically different from zero (F-value: 27.90).
  *   We re-estimated the model with the pooled OLS estimator, using clustered standard errors.

## Example: Pooled OLS (Cont.)

- This is what the data look like:

```{r}
#| echo: false
#| warning: false
#| message: false
library(xtsum)
library(wooldridge); library(modelsummary); library(tinytable)

# Load the wagepan dataset
data("wagepan")

xtsum(data=wagepan, id = 'nr', t= 'year', variables= c('union', 'black', 'educ', 'hisp', 'exper', 'expersq'))
```


## Example: Pooled OLS (Cont.)

- Results show that unions raise wages by about 18%

```{r}
#| echo: false
#| warning: false
#| message: false

# Install and load necessary packages
library(wooldridge); library(lmtest); library(fixest); library(modelsummary); library(tinytable)

# Load the wagepan dataset
data("wagepan")

# 1. Pooled OLS Regression
pooled_ols <- lm(lwage ~ union + educ + black + hisp + exper + expersq + married, data=wagepan)

# 3. Pooled OLS with Clustered Standard Errors
pooled_ols_cluster <- feols(lwage ~ educ + black + hisp + exper + expersq + married + union, data=wagepan, vcov=~as.factor(nr))

modelsummary(list('Pooled OLS'=pooled_ols, 'Pooled OLS, Clustered SE'=pooled_ols_cluster), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example: Pooled OLS (Cont.)

- Since there is strong evidence for autocorrelation according to the BG-test, cluster standard errors by individual. 

```{r}
#| echo: false
#| warning: false
#| message: false
# 2. Breusch-Godfrey Test for Autocorrelation
# The null hypothesis is that there is no serial correlation.
uhat <- residuals(pooled_ols)
uhat_lag1 <- dplyr::lag(uhat, 1)
diff_regressors <- model.matrix(pooled_ols)[, -1] # We drop the intercept
aux_data <- na.omit(data.frame(uhat, uhat_lag1, diff_regressors))
aux_reg <- lm(uhat ~ ., data = aux_data)
modelsummary(list('Breusch-Godfrey Test'=aux_reg),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Pooled OLS Visualization

```{r}
#| echo: false
#| fig-align: 'center'
#| fig-width: 10
# Load the necessary libraries
library(ggplot2)
library(dplyr)

#--- 1. Simulate Panel Data ---

# Set a seed for reproducibility of the random data
set.seed(123)

# Define the parameters for the data simulation
n_countries <- 5      # Number of countries
n_years <- 20         # Number of years per country
total_obs <- n_countries * n_years

# Create country-specific intercepts to simulate unobserved heterogeneity
# In a real-world scenario, these are the unique, unobserved effects.
country_effects <- runif(n_countries, min = -20, max = 20)

# Create the panel data frame
panel_data <- data.frame(
  country = factor(rep(1:n_countries, each = n_years, labels = paste("Country", 1:n_countries))),
  year = rep(1:n_years, times = n_countries)
) %>%
  # Create the independent variable (X) and the dependent variable (Y)
  mutate(
    # X is an independent variable that grows over time with some noise
    x = year + rnorm(total_obs, mean = 0, sd = 2),
    
    # Y depends on X, a country-specific effect, and random noise
    # The true relationship has a slope of 2.5
    y = 2.5 * x + rep(country_effects, each = n_years) + rnorm(total_obs, mean = 0, sd = 10)
  )

#--- 2. Perform Pooled OLS Estimation ---

# Run a simple linear regression, ignoring the panel structure (country and year)
# This is the "pooled" OLS model.
pooled_model <- lm(y ~ x, data = panel_data)

#--- 3. Create the ggplot Visualization ---

ggplot(panel_data, aes(x = x, y = y, color = country)) +
  # Add the scatter plot points, colored by country
  geom_point(alpha = 0.7, size = 2) +
  
  # Add the single regression line from the pooled OLS model
  # geom_smooth() with method = "lm" calculates and plots this line for the entire dataset
  geom_smooth(method = "lm", se = FALSE, color = "black", aes(group = 1), linetype = "dashed") +
  
  # Add labels and a title for clarity
  labs(
    title = "Pooled OLS Estimation on Panel Data",
    subtitle = "A Single Regression Line Fitted to All Observations",
    x = "Independent Variable (X)",
    y = "Dependent Variable (Y)",
    color = "Country"
  ) +
  
  # Use a clean and minimal theme
  theme_minimal() +
  
  # Improve the legend and overall appearance
  theme(
    legend.position = "bottom",
    plot.title = element_text(hjust = 0.5, face = "bold"),
    plot.subtitle = element_text(hjust = 0.5)
  )
```

# Random Effects

## Random Effects Model

- Consider the static model:

  $$
    y_{it} = \beta_1x_{1it} + \dots + \beta_kx_{kit} + a_i + u_{it} \quad (6)
  $$

- for $i=1,...,N; t=1,...,T$

-   The equation (6) does not allow correlation between $a_i$ and all of the right hand side variables $x_{1it},...,x_{kit}$ (over all $T$ periods)
    $$
      E(a_i | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}) = 0
    $$
-   The exogenous time-varying regressors are assumed to be strictly exogenous (conditional on the unobserved effect):
    $$
      E(u_{it} | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}, a_i) = 0
    $$
-   In simple words: there are neither lagged dependent variables nor is there any feedback mechanism

## Random Effects Model (Cont.)

-   The following is assumed about $u_{it}$:
    -   $E(u_{it}) = 0$; $\text{Var}(u_{it}) = \sigma_u^2$ (expected value of zero; constant variance)
    -   $u_{it}$ is independent over time and across individuals
-   The following is assumed about $a_i$:
    -   $E(a_i) = 0$; $\text{Var}(a_i) = \sigma_a^2$ (expected value of zero; constant variance)
    -   $a_i$ is independent over time and across individuals
-   Equation (6) can be rewritten as follows:
    Model: $y_{it} = \beta_1x_{1it} + \dots + \beta_kx_{kit} + v_{it} \quad (6)$
    $i=1,...,N; t=1,...,T$
    where
    $$v_{it} = a_i + u_{it} \quad (7)$$

-   The random effects estimator works as follows. We estimate equation (6), in which the error structure of equation (7) is taken into account.


## Correlation Structure Error Terms

-   Error term (see slide above): $v_{it} = a_i + u_{it}$
-   Covariance across two individuals $i$ and $j$:
    -   $\text{Cov}(v_{it}, v_{jt}) = 0$ for $i \neq j$
    -   $\text{Cov}(v_{it}, v_{js}) = 0$ for $i \neq j; t \neq s$
-   Covariance for the same individual across time:
    -   $\text{Cov}(aX+bY, cW+dZ) = ac\text{Cov}(X,W) + ad\text{Cov}(X,Z) + bc\text{Cov}(Y,W) + bd\text{Cov}(Y,Z)$
-   So that:
    $$
      \begin{align*}
      \text{Cov}(v_{it}, v_{is}) &= \text{Cov}(a_i + u_{it}, a_i + u_{is}) \\
      &= \text{Cov}(a_i, a_i) + \text{Cov}(a_i, u_{is}) + \text{Cov}(u_{it}, a_i) + \text{Cov}(u_{it}, u_{is}) \\
      &= \sigma_a^2 + 0 + 0 + 0 \\
      &= \sigma_a^2
      \end{align*}
    $$

## Correlation Structure Error Terms

-   Variance for an individual:
    $$\text{Var}(X+Y) = \text{Var}(X) + \text{Var}(Y) + 2\text{Cov}(X,Y)$$
    $$
    \begin{align*}
    \text{Var}(v_{it}) &= \text{Var}(a_i + u_{it}) \\
    &= \text{Var}(a_i) + \text{Var}(u_{it}) + 2\text{Cov}(a_i, u_{it}) \\
    &= \sigma_a^2 + \sigma_u^2 + 0
    \end{align*}
    $$
-   Thus:
    $$\text{Corr}(v_{it}, v_{is}) = \frac{\text{Cov}(v_{it}, v_{is})}{\sqrt{\text{Var}(v_{it})\text{Var}(v_{is})}} = \frac{\sigma_a^2}{\sigma_a^2 + \sigma_u^2}$$
-   Conclusion: random effects estimator $\beta_{re}$ takes account of the autocorrelation structure

## Random Effects: Further Remarks

- Both random effects and pooled OLS allow for the inclusion of time-invariant individual variables (e.g. gender in a wage equation):

  $$
    y_{it} = a_i + \beta_1x_{1it} + \dots + \beta_kx_{kit} + \gamma_1z_{1i} + \dots + \gamma_rz_{ri} + u_{it}
  $$
- For $i=1,...,N; \quad t=1,...,T$. 
- And $z$ is a vector of individual-specific regressors.
- Remember: the effect of $z$ on $y$ cannot be estimated ('is not identified') in the fixed-effects specification.

## Estimation of RE Models (GLS)

- Estimating the RE model can "almost" be done using Pooled OLS. 
  - If we run Pooled OLS on the data, assuming the RE model, the estimates of $\beta$ will be unbiased (if the key assumption holds), but they will be **inefficient**.

- The composite error term $u_{it} = \alpha_i + \nu_{it}$ creates serial correlation within each individual. 
- **Solution:** Generalized Least Squares (GLS).
    - GLS is a method that transforms the data to account for this specific error structure, producing efficient estimates.
    - In practice, we use **Feasible GLS (FGLS)** because we have to estimate the components of the error correlation first. This is what statistical software does automatically.

## Interpretation of RE Coefficients

- The RE estimator uses a weighted average of the "within" and "between" variation in the data.

:::{.callout-tip title="Interpretation of RE Estimates"}

The coefficient $\beta_1$ from an RE model is interpreted as:

The estimated change in $y$ for a one-unit increase in $x$, assuming the unobserved individual effects $\alpha_i$ are uncorrelated with $x$.

:::

- It's a more general interpretation than in the FE model.
- The reliability of this interpretation hinges entirely on the key RE assumption holding true.

## Pros and Cons of the Random Effects Model

- **Pros:**
    - **Can estimate the effects of time-invariant variables** (e.g., gender, race, industry), because it does not wipe them out.
    - **More efficient** (i.e., has smaller standard errors) than the FE model, *if* the key assumption ($E(\alpha_i | X_{it}) = 0$) is met. It uses both "within" and "between" variation.

- **Cons:**
    - **Estimates are biased and inconsistent if the key assumption is violated.** This is the critical weakness. If the unobserved effects are correlated with your regressors, the RE model suffers from omitted variable bias.


# Hausman Test

## RE or FE? 

- Model: $y_{it} = \beta_1x_{1it} + \dots + \beta_kx_{kit} + a_i + u_{it} \quad (8)$
  - $E(a_i | x_{1i1},\dots,x_{1iT},\dots,x_{ki1},\dots,x_{kiT}) = 0$
  - $E(u_{it} | x_{1i1},\dots,x_{1iT},\dots,x_{ki1},\dots,x_{kiT}, a_i) = 0$

- $x_{it}$ is strictly exogenous (in simple words: no lagged dependent variables; no feedback mechanism).

- **Estimation method: Random effects**
  - Equation (8) can be estimated with random effects, which requires the assumption $E(a_i | x_{1i1},\dots,x_{1iT},\dots,x_{ki1},\dots,x_{kiT}) = 0$
  - $\hat{\beta}_{re}$ is a consistent estimator of $\beta$

- **Estimation method: Fixed effects**
  - Equation (8) can be estimated with fixed effects, which does NOT assume that $E(a_i | x_{1i1},\dots,x_{1iT},\dots,x_{ki1},\dots,x_{kiT}) = 0$
  - $\hat{\beta}_{within}$ is a consistent estimator of $\beta$
  - $\hat{\beta}_{within}$ is less efficient than $\hat{\beta}_{re}$: It means that $\text{Var}(\hat{\beta}_{re}) < \text{Var}(\hat{\beta}_{within})$
- Consequently, random effects yields significant t-statistics more easily than fixed effects.

## Comparing FE and RE: The Hausman Test

- So which model should we use? FE or RE? The **Hausman Test** helps us decide.

- **Intuition:**
    - The FE estimator is **always consistent**, whether the unobserved effects $\alpha_i$ are correlated with $X$ or not.
    - The RE estimator is **consistent AND efficient** if $\alpha_i$ and $X$ are uncorrelated, but **inconsistent** if they are correlated.

- **The Test:** We compare the coefficient estimates from FE and RE.
    - If the coefficients are "close" to each other, it suggests the RE assumption holds, and we should use the more efficient RE model.
    - If the coefficients are "far apart" and statistically different, it suggests the RE assumption is violated. We must use the consistent FE model.
    
## Hausman Test Structure

- Model: $y_{it} = \beta_{1}x_{1it} + ... + \beta_{k}x_{kit} + a_{i} + u_{it}$ 

  - $E(a_{i} | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}) = 0$

  - $E(u_{it} | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}, a_{i}) = 0$

- **Structure:**
  - Null hypothesis: $H_{0}: E(a_{i} | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}) = 0$
  - Alternative hypothesis: $H_{1}: E(a_{i} | x_{1i1},...,x_{1iT},...,x_{ki1},...,x_{kiT}) \neq 0$

- Under the null hypothesis ($H_{0}$), random effects is preferred (because of the zero correlation between the $a_{i}$ and the explanatory variables). 
  - The model can be estimated by random effects and fixed effects. 
  - If $H_{0}$ is true, both estimators are unbiased (consistent). However, random effects yields smaller standard errors, so that it is preferred to fixed effects.
- If alternative hypothesis ($H_{1}$) is true: fixed effects is preferred, because random effects estimator is biased. 

## Hausman Test Statistic

- The test is based on the difference between the coefficient vectors from the two models: $(\hat{\beta}_{FE} - \hat{\beta}_{RE})$.

- The Hausman statistic, $H$, is a measure of the squared distance between the two vectors of coefficients, weighted by the precision (1/variance) of this difference. 
  - It is constructed so that large, systematic differences between the coefficients lead to a large test statistic.

- Under the null hypothesis ($H_0$), the test statistic follows a Chi-squared distribution with $k$ degrees of freedom, where $k$ is the number of time-varying regressors in the model.

$$ H \sim \chi^2(k) $$

## $\chi^2$ Test Visualization

- The $\chi^2$ distribution is one-sided.
  - A significance level $\alpha$ gives you a critical value on the basis of which a test can be rejected. 
  - Alternatively, the $p$-value can be calculated according to the cdf. 
  
```{r}
#| echo: false
#| fig-align: 'center'
#| fig-width: 10
#| fig-height: 4
# Load required packages
library(ggplot2)

# Set degrees of freedom for the Chi-squared distribution
df <- 5

# Generate Chi-squared distribution data
x <- seq(0, 20, length.out = 1000)
chi_sq_density <- dchisq(x, df)

# Create a data frame for plotting
plot_data <- data.frame(x = x, density = chi_sq_density)

# Generate a random critical value (greater than median for demonstration)
set.seed(123)  # For reproducibility
critical_value <- runif(1, qchisq(0.5, df), qchisq(0.99, df))

# Calculate the p-value (right-tailed test)
p_value <- pchisq(critical_value, df, lower.tail = FALSE)

# Find the y-position for the critical value label (just above the line)
label_y <- dchisq(critical_value, df) * 1.1

# Find the x-position for the p-value label (in the middle of the tail area)
p_label_x <- critical_value + (max(x) - critical_value)/2
p_label_y <- max(chi_sq_density)/4  # Position in the empty space

# Plot the Chi-squared distribution
p <- ggplot(plot_data, aes(x = x, y = density)) +
  geom_line(color = "blue", linewidth = 1) +
  geom_area(data = subset(plot_data, x > critical_value), 
            aes(y = density), fill = "red", alpha = 0.3) +
  geom_vline(xintercept = critical_value, color = "darkgreen", linetype = "dashed") +
  
  # Critical value label positioned at the line
  geom_text(aes(x = critical_value, y = label_y, 
                label = paste("Critical value =", round(critical_value, 2))),
            hjust = -0.1, vjust = 0, color = "darkgreen") +
  
  # p-value label positioned in the tail area
  geom_text(aes(x = p_label_x, y = p_label_y, 
                label = paste("p-value =", round(p_value, 4))),
            color = "darkred") +
  
  labs(title = paste("Chi-squared Distribution with", df, "Degrees of Freedom"),
       x = "Value", y = "Density") +
  theme_minimal()

p
```

## The Hausman Test: Application

:::{.callout-note title="Hausman Test: Procedure"}
- $H_0$: The Random Effects model is the appropriate model. (The difference in coefficients between FE and RE is not systematic, i.e., $E(\alpha_i | X_{it}) = 0$).

- $H_A$: The Fixed Effects model is the appropriate model. (The difference in coefficients is systematic, i.e., $E(\alpha_i | X_{it}) \neq 0$).

- Statistical software calculates a test statistic (Chi-squared) and a p-value.
    - **If p-value < 0.05 (or your chosen significance level):** Reject the null hypothesis. The models are significantly different. Conclude that the RE assumption is likely violated. **Use the Fixed Effects model.**
    - **If p-value >= 0.05:** Fail to reject the null hypothesis. You do not have evidence that the RE assumption is violated. **Use the more efficient Random Effects model.**

:::

## Example: FE vs. RE 

- We use the same dataset as before (Vella and Verbeek, 1998)
- We estimate the model: $\log \text{Wage} = \alpha_i + \beta_1 \text{Union}_{it} + \beta_2 \text{Exper}_{it} + \beta_3 \text{Exper}^2_{it} + u_{it}$ by FE and RE
  - The Hausman test can only compare models with the same variables, hence we only include time-varying control variables. 
- The results are shown below:

```{r}
#| echo: false
#| warning: false
#| message: false
# Install and load necessary packages if you haven't already.
# install.packages(c("wooldridge", "plm"))
library(wooldridge); library(plm); library(modelsummary)

# Load the wagepan dataset
data("wagepan")

# Define the model formula
# We will model the log of wage (lwage) as a function of several explanatory variables.
# Note: Time-invariant variables like 'educ', 'black', and 'hisp' will be
# dropped from the Fixed Effects model automatically.
formula <- lwage ~ union + exper + expersq

# 1. Estimate the Random Effects (RE) Model
# The RE model assumes that the unobserved individual effects (a_i) are
# uncorrelated with the explanatory variables.
re_model <- plm(formula,
                data = wagepan,
                index = c("nr", "year"),
                model = "random")


# 2. Estimate the Fixed Effects (FE) Model
# The FE model (or 'within' estimator) allows the unobserved individual effects (a_i)
# to be correlated with the explanatory variables. It removes the a_i by
# demeaning the variables.
fe_model <- plm(formula,
                data = wagepan,
                index = c("nr", "year"),
                model = "within")

modelsummary(list('FE'=fe_model, 'RE'=re_model), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example: Hausman Test

- According to these models, union membership raises wages with 8.3% (FE) or 10.2% (RE). 
- The Hausman test clearly rejects the null hypothesis of $E(\alpha_i | X_{it}) = 0$. 
- This implies the fixed effects model is preferable to the random effects model.

```{r}
#| echo: false
# 3. Conduct the Hausman Test
# The Hausman test compares the FE and RE models.
# H0 (Null Hypothesis): The Random Effects model is appropriate (i.e., individual effects are not correlated with regressors).
# H1 (Alternative Hypothesis): The Fixed Effects model is appropriate (i.e., individual effects are correlated with regressors).
hausman_test_result <- phtest(fe_model, re_model)

# Print the results of the Hausman test
print(hausman_test_result)
```

# Pooled OLS vs. Random Effects

## Pooled OLS vs. RE

*   To formally decide between Pooled OLS and Random Effects, you should use the Breusch-Godfrey test for random effects.

*   **Purpose of the Test:** This test checks whether the variance of the individual-specific effect ($a_i$) is equal to zero.
*   **Hypotheses:**
    *   **Null Hypothesis ($H_0$):** $Var(a_i) = \sigma_a^2 = 0$. There are no significant individual-specific effects. The Pooled OLS model is adequate.
    *   **Alternative Hypothesis ($H_1$):** $Var(a_i) = \sigma_a^2 > 0$. There are significant individual-specific effects. The Random Effects model is preferred over Pooled OLS.
*   **Decision Rule:**
    *   If the **p-value is high** (e.g., > 0.05), you fail to reject the null hypothesis. This suggests that the individual effects are not significant, and you can use the simpler **Pooled OLS** model.
    *   If the **p-value is low** (e.g., < 0.05), you reject the null hypothesis. This provides strong evidence for the presence of individual-specific effects, making the **Random Effects** model the better choice.

## Example: Pooled OLS vs. RE

- Using again the same dataset (Vella and Verbeek, 1998)
  - We can now use the full set of control variables. 
  - The results are shown below.

```{r}
#| echo: false
#| message: false
#| warning: false
# Install and load necessary packages
library(wooldridge)
library(plm); library(modelsummary)

# Load the dataset
data("wagepan")

# Define the model formula
formula <- lwage ~ union + educ + black + hisp + exper + expersq + married

# 1. Estimate a Pooled OLS model
pooled_ols <- plm(formula, data = wagepan, model = "pooling")
re <- plm(formula, data = wagepan, model = "random")


modelsummary(list('Pooled OLS'=pooled_ols, 'Random Effects'=re), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```


## Example: Pooled OLS vs. RE (Cont.)

:::{style="font-size:0.8em;"}
- The results of the Breusch-Godfrey test show the variance of the unobserved effects $\text{Var}(\sigma^2_{a}) > 0$.
  - This implies the Random Effects model is more reliable. 
  - Hence the Random Effects model should be preferred. 

```{r}
#| echo: false
#| warning: false
#| message: false
# The null hypothesis is that there is no serial correlation.
uhat <- residuals(pooled_ols)
uhat_lag1 <- dplyr::lag(uhat, 1)
diff_regressors <- model.matrix(pooled_ols)[, -1] # We drop the intercept
aux_data <- na.omit(data.frame(uhat, uhat_lag1, diff_regressors))
aux_reg <- lm(uhat ~ ., data = aux_data)
modelsummary(list('Breusch-Godfrey Test'=aux_reg),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_latex(resize_direction = 'down', resize_width = 0.2)
```

:::

# Choosing Panel Models

## Reasoning with Panel Data

:::{.callout-tip title="Line of Reasoning with Panel Data"}

-   **Step 1a:** Apply pooled OLS
    $$y_{it} = \beta_1x_{1it} + \dots + \beta_kx_{kit} + v_{it}$$
-   **Step 1b:** Test for autocorrelation of error term with Breusch-Godfrey. Estimated parameter on lagged $v_{it}$ is
    $$\text{Corr}(v_{it}, v_{is}) = \frac{\sigma_a^2}{\sigma_a^2 + \sigma_u^2}$$
-   **Step 1c:** Re-estimate pooled OLS with clustered standard errors
-   **Step 2a:** Apply first-differences estimator (FD):
    $$\Delta y_{it} = \beta_1\Delta x_{1it} + \dots + \beta_k\Delta x_{kit} + \Delta u_{it}$$
-   **Step 2b:** Compute autocorrelation: is the parameter on $\Delta u_{i,t-1}$ of Breusch-Godfrey equal to -0.5?
-   **Step 2c:** Re-estimate FD with clustered standard errors
- (Continued on next slide.)

:::


## Reasoning with Panel Data (Cont.)

:::{.callout-tip title="Line of Reasoning with Panel Data"}

-   **Step 3a:** Apply LSDV/ fixed-effects estimator (FE):
    $$\ddot{y}_{it} = \beta_1\ddot{x}_{1it} + \dots + \beta_k\ddot{x}_{kit} + \ddot{u}_{it}$$
-   **Step 3b:** Compute autocorrelation
-   **Step 3c:** Compare the outcome of the FD estimator with the outcome of the FE estimator
-   **Step 4a:** Apply random effects estimator (RE)
-   **Step 4b:** Test for autocorrelation
-   **Step 4c:** Compare the outcome of the Random effects estimator and Pooled OLS estimator
-   **Step 5:** Test for Random effects versus fixed effects using the Hausman test.
:::


## Example (Entire Procedure)

- We apply the five steps on the Vella and Verbeek (1998) dataset
- Step 1a: Apply Pooled OLS: 

```{r pooled_ols}
modelsummary(list('Pooled OLS'=pooled_ols), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```


## Example (Entire Procedure) (Cont.)

- Step 1b: Test for autocorrelation of error term with Breusch-Godfrey.

```{r}
#| echo: false
library(lmtest); library(plm)
# 2. Breusch-Godfrey Test for Autocorrelation
# The null hypothesis is that there is no serial correlation.
uhat <- residuals(pooled_ols)
uhat_lag1 <- dplyr::lag(uhat, 1)
diff_regressors <- model.matrix(pooled_ols)[, -1] # We drop the intercept
aux_data <- na.omit(data.frame(uhat, uhat_lag1, diff_regressors))
aux_reg <- lm(uhat ~ ., data = aux_data)
modelsummary(list('Breusch-Godfrey Test'=aux_reg),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example (Entire Procedure) (Cont.)

- Step 1c: Re-estimate pooled OLS with clustered standard errors


```{r pooled_clust}
#| echo: false
library(fixest)
# Define the model formula
formula <- lwage ~ union + educ + black + hisp + exper + expersq + married

# 1. Estimate a Pooled OLS model
pooled_ols <- feols(formula, data = wagepan, vcov=~nr)
modelsummary(list('Pooled OLS, Clustered SE'=pooled_ols), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example (Entire Procedure) (Cont.)

- Step 2a: Apply first-differences estimator (FD)
- Only time-varying variables remain in the equation.

```{r first_diff}
#| echo: false
fd <- plm(formula, data=wagepan, index=c("nr", "year"), model="fd")
modelsummary(list('FD'=fd), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example (Entire Procedure) (Cont.)

- Step 2b: Compute autocorrelation: is the parameter on $\Delta u_{it-1}$ of Breusch-Godfrey equal to -0.5?
- $-0.5$ is not in the confidence interval. Therefore we prefer FD over the FE estimator (see previous lecture).

```{r}
#| echo: false
library(dplyr)
uhat <- residuals(fd)
uhat_lag1 <- dplyr::lag(uhat, 1)
diff_regressors <- model.matrix(fd)[, -1] # We drop the intercept
aux_data <- na.omit(data.frame(uhat, uhat_lag1, diff_regressors))
aux_reg <- lm(uhat ~ ., data = aux_data)
modelsummary(list('Breusch-Godfrey Test'=aux_reg),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```


## Example (Entire Procedure) (Cont.)

- Step 2c: Re-estimate FD with clustered standard errors

```{r}
#| echo: false
fd <- plm(formula, data=wagepan, index=c("nr", "year"), model="fd")
fd$vcov <- vcovHC(fd, method="arellano", cluster="group")
modelsummary(list('FD, Clustered SE'=fd),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```


## Example (Entire Procedure) (Cont.)

- Step 3a: Apply LSDV/fixed-effects estimator (FE)

```{r}
#| echo: false
library(fixest)
formula2 <- lwage ~ union + educ + black + hisp + exper + expersq + married | nr
fe <- feols(formula2, data=wagepan)
modelsummary(list('FE'=fe),
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```


## Example (Entire Procedure) (Cont.)

- Step 4a: Apply random effects estimator

```{r}
#| echo: false
# Define the model formula
formula <- lwage ~ union + educ + black + hisp + exper + expersq + married

# 1. Estimate a Pooled OLS model
re <- plm(formula, data = wagepan, model = "random")


modelsummary(list('Random Effects'=re), 
             stars=c("*"=0.1, "**"=0.05, "***"=0.10),
             width=c(0.2, 0.15),
             gof_map=c('r.squared', 'nobs')) |>
  theme_revealjs(fontsize=1)
```

## Example (Entire Procedure) (Cont.)

- Step 5: Test for Random effects versus fixed effects using Hausman test.
  - The Hausman test convincingly rejects the null hypothesis, hence Fixed Effects is preferred over Random Effects. 

```{r}
#| echo: false
fe <- plm(formula, data = wagepan, model = "within")
phtest(re, fe)
```


# Conclusion

## Conclusion 

*   This lecture expanded our toolkit for panel data by introducing **Pooled OLS** and the **Random Effects (RE)** estimator, providing alternatives to the Fixed Effects (FE) and First Differences models. 

*   The critical choice between Fixed Effects (FE) and Random Effects (RE) hinges on a key assumption: Is the unobserved individual effect ($a_i$) correlated with the explanatory variables ($X_{it}$)?

*   **Model Summary:**
    *   **Use Fixed Effects (FE)** if you believe $a_i$ and $X_{it}$ are correlated. This provides **consistent** estimates, but you cannot estimate the effect of time-invariant variables.
    *   **Use Random Effects (RE)** if you can assume $a_i$ and $X_{it}$ are uncorrelated. This model is more **efficient** (has smaller standard errors) and, importantly, allows you to estimate the effects of time-invariant variables.
    *   **Pooled OLS** is the simplest model but is often inappropriate due to its strict assumption that there are no unique individual effects, which can lead to omitted variable bias.


## Conclusion (Cont.)

*   **Decision Framework:**
    *   The **Breusch-Godfrey test** helps you choose between Pooled OLS and Random Effects by testing for the presence of significant individual effects.
    *   The **Hausman test** is the crucial tool for deciding between the Fixed Effects and Random Effects models. It formally tests whether the individual effects are correlated with the regressors.
        *   A significant result (low p-value) favors the **Fixed Effects model**.
        *   An insignificant result favors the more efficient **Random Effects model**.

*   We have established a systematic line of reasoning: estimate the various models (Pooled OLS, FE, RE) and use this series of statistical tests to justify your final model choice.


# The End

